---
title: "Weather Events Causing Most Damage"
author: "Guru Medasani"
date: "October 18, 2015"
output: html_document
---

## Synopsis

In this report we analyze the storm database from National Weather Service. 
Events in database start in the year 1950 and end in November 2011. We explore
the dataset to find the top 5 weather event types that caused highest number of
fatalities and injuries. We also explore the dataset to find the top 5 weather
even types that caused greatest economic damage.

Note: To learn more about the dataset, please visit https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf

## Data Processing

### Load the libraries
```{r}
library(ggplot2)
library(dplyr)
library(knitr)
library(scales)
library(xtable)
```

### Load the dataset and explore the dataset
```{r, cache=TRUE}
fileName <- ("/Users/gmedasani/Documents/courses/Coursera/DataScience-Track/
ReproducibleResearch/Week3/project2/repdata%2Fdata%2FStormData.csv.bz2")
fileName <- gsub("\n", "", fileName)
storm_data <- read.csv(fileName)
dim(storm_data)
names(storm_data)
```

### Clean the dataset

Based on the Questions 1 and 2, we are only intested in the following 
variables of the storm database

* EVTYPE
* FATALITIES
* INJURIES
* PROPDMG
* PROPDMGEXP
* CROPDMG
* CROPDMGEXP

```{r, cache=TRUE}
storm_data <- select(storm_data, EVTYPE,FATALITIES,INJURIES,
                           PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP)
```

### Question 1

#### Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

```{r kable, cache=TRUE}

storm_fatalities_injuries_by_evtype <- group_by(select(storm_data, 
                                              EVTYPE,FATALITIES,INJURIES), 
                                       EVTYPE)

storm_affected_population <- summarise(storm_fatalities_injuries_by_evtype,
                             total_fatalities = sum(FATALITIES, na.rm=TRUE),
                             total_injuries = sum(INJURIES, na.rm=TRUE))

storm_affected_population <- select(mutate(storm_affected_population, 
                affected_population = total_fatalities +
                total_injuries), weather_event_type = 
                        EVTYPE,affected_population)

storm_affected_population_ordered <- arrange(storm_affected_population, 
                                             affected_population)

top5_evtypes_by_affected_population <- tail(storm_affected_population_ordered
                                             ,5)

top5_evtypes_by_affected_population_formatted <- mutate(
        top5_evtypes_by_affected_population,
                                        affected_population = 
                                        comma(affected_population))

```

### Question 2

#### Across the United States, which types of events have the greatest economic consequences?

```{r cache=TRUE}
findMagnitude <- function(magnitudeCharacter){
        magnitudeCharacter <- toupper(as.character(magnitudeCharacter))
        multiplyNumber <- 0
        if ( magnitudeCharacter == 'B'){
                multiplyNumber <- 1000000000
        }else if (magnitudeCharacter == 'M'){
                multiplyNumber <- 1000000
        }else if (magnitudeCharacter == 'K'){
                multiplyNumber <- 1000
        }else if (magnitudeCharacter == 'H'){
                multiplyNumber <- 100
        }else{
                multiplyNumber <- 1
        }
        return(multiplyNumber)
}

storm_data['PROPDMGEXP'] <- sapply(storm_data$PROPDMGEXP, 
                                             findMagnitude)
storm_data['CROPDMGEXP'] <- sapply(storm_data$CROPDMGEXP,
                                             findMagnitude)
storm_data <- select(mutate(storm_data, 
                               PROPDMG = as.numeric(PROPDMG*PROPDMGEXP),
                               CROPDMG = as.numeric(CROPDMG*CROPDMGEXP),
                               damage_in_dollars = PROPDMG + CROPDMG),
                               weather_event_type = EVTYPE,
                               damage_in_dollars)
storm_damage_by_evtype <- summarise(group_by(storm_data,weather_event_type), 
                                total_damage_in_dollars = 
                                sum(damage_in_dollars, na.rm = TRUE))

storm_damage_by_evtype_ordered <- arrange(storm_damage_by_evtype,
                                          total_damage_in_dollars)

top5_evtypes_by_economic_damage <- tail(storm_damage_by_evtype_ordered,5)

top5_evtypes_by_economic_damage_formatted <- 
        mutate(top5_evtypes_by_economic_damage,
        total_damage_in_dollars = 
        dollar(total_damage_in_dollars))
```

## Results

### Top 5 Weather Events by Affected Population

```{r, fig.width=15, fig.height=7.5, results='asis'}
p1 <- ggplot(data = top5_evtypes_by_affected_population, 
             aes(x = affected_population, y = weather_event_type))

plot1 <- p1 + 
geom_point(aes(size = top5_evtypes_by_affected_population$affected_population,
        colour = top5_evtypes_by_affected_population$weather_event_type,
        label = top5_evtypes_by_affected_population$affected_population)) + 
geom_text(aes(label = 
        top5_evtypes_by_affected_population_formatted$affected_population), 
        size = 4) +
scale_size(range = c(10,25)) +
scale_y_discrete(limits = 
         top5_evtypes_by_affected_population$weather_event_type)+
theme_bw()

print(xtable(top5_evtypes_by_affected_population_formatted),type="html")
```

### Fig 1: Top 5 Weather Events by Affected Population
```{r, fig.width=15, fig.height=7.5 }
plot1
```

### Top 5 Weather Events by Economic Damage (in Dollars)

```{r,fig.width=17, fig.height=7.5, results='asis'}
p2 <- ggplot(data = top5_evtypes_by_economic_damage, 
             aes(x = total_damage_in_dollars, y = weather_event_type))

plot2 <- p2 + 
geom_point(aes(size = top5_evtypes_by_economic_damage$total_damage_in_dollars,
           colour = top5_evtypes_by_economic_damage$weather_event_type,
           label = top5_evtypes_by_economic_damage$total_damage_in_dollars)) + 
geom_text(aes(label = 
        top5_evtypes_by_economic_damage_formatted$total_damage_in_dollars), 
        size = 4) +
scale_size_continuous(range = c(15,30)) + 
scale_y_discrete(limits = top5_evtypes_by_economic_damage$weather_event_type)+
theme_bw()

print(xtable(top5_evtypes_by_economic_damage_formatted),type="html")
```

### Fig 2: Top 5 Weather Events by Economic Damage (in Dollars)

```{r, fig.width=17, fig.height=7.5}
plot2
```






